---
title: "The Lunar Renaissance"
format: html
---


## Lunar missions

The story of the moon's abandonment, and the arrival of new players. To accompany my essay on returning to the Moon.

Data sourced from Wikipedia: <https://en.wikipedia.org/wiki/List_of_missions_to_the_Moon>

For similar visualisations, see:

-   <https://www.lpi.usra.edu/lunar/missions/>
-   <https://www.visualcapitalist.com/cp/visualizing-all-attempted-and-successful-moon-landings/>
-   <https://www.scientificamerican.com/article/50-years-of-moon-missions-graphic/>)

```{r}
#| label: libraries
#| include: false

# Install and load the necessary packages
library(rvest)
library(tidyverse)
library(janitor)
library(lubridate)
library(plotly)
library(knitr)
library(plotly)
```

```{r}
#| label: importing data

# Specify the URL of the Wikipedia page
url <- "https://en.wikipedia.org/wiki/List_of_missions_to_the_Moon"

# Read the page
page <- read_html(url)

# Extract the first table with the class 'wikitable'
tables <- page %>%
  html_nodes("table.wikitable") %>%
  html_table(fill = TRUE)

# Define function to clean raw tables
clean_table <- function(df_raw) {
  df_raw <- df_raw[, -1] |> clean_names() # Remove index columns and clean column names
  # Remove any footnotes
  df_raw <- as.data.frame(lapply(df_raw, function(x)
    str_remove_all(x, "\\[\\d+\\]")))
  # Move description rows to new column
  df <- filter(df_raw, spacecraft != launch_date)
  mission_descriptions <- filter(df_raw, spacecraft == launch_date) |> rename(description = spacecraft) |>
    select(1:2)
  df <- merge(df, mission_descriptions)
  
  return(df)
}

c20_clean <- clean_table(tables[[1]])
c21_clean <- clean_table(tables[[2]])

moon_missions <- rbind(c20_clean, c21_clean)
```

```{r}
#| label: clean-data

# Format dates
moon_missions$launch_date <- lubridate::dmy(moon_missions$launch_date)
moon_missions$year <- lubridate::year(moon_missions$launch_date)

# Fixing description
moon_missions$description[moon_missions$spacecraft=="Ranger 5 lander"] <- "Solar panels erroneously disengaged from power system, failed 9 hours after launch when batteries were depleted. Missed the Moon as course correction was not completed."

# Get country names
agency_countries <- list(
  "NASA" = "USA",
  "CNSA" = "China",
  "JAXA" = "Japan",
  "ESA" = "Europe",
  "ISRO" = "India",
  "Lavochkin" = "USSR/Russia",
  "OKB-1" = "USSR/Russia",
  "Roscosmos" = "USSR/Russia"
)

# Replace using gsub with fixed = TRUE to match exact strings
moon_missions <- moon_missions |> 
  mutate(country = case_when(
    operator == "NASA" ~ "USA",
    operator == "CNSA" ~ "China",
    operator == "JAXA" ~ "Japan",
    operator == "ESA" ~ "Europe",
    operator == "ISRO" ~ "India",
    operator == "Lavochkin" ~ "USSR/Russia",
    operator == "OKB-1" ~ "USSR/Russia",
    operator == "Roscosmos" ~ "USSR/Russia",
    TRUE ~ operator))


# Creating concise "spacecraft (mission)" name

moon_missions <- moon_missions |> 
  rowwise() |> 
  mutate(mission_new = ifelse(
    grepl(spacecraft, mission),
    paste0("<b>",spacecraft,"</b>"),
    paste0("<b>",spacecraft,"</b>", " (", mission, ")")
  ))

# Simplifying statuses

moon_missions <- moon_missions %>%
  mutate(
    outcome_simple = case_when(
      outcome %in% c("Operational", "Success", "En route") ~ "Success",
      .default = "Failure"
    )
  )

# Putting line breaks in description
moon_missions <- moon_missions %>%
  mutate(description = str_wrap(description, width = 50)) %>%
  mutate(description = str_replace_all(description, "\n", "<br>"))

```

```{r}
#| label: decade-table

decade_table <- moon_missions %>%
  mutate(decade = floor(year(launch_date) / 10) * 10) |> 
  group_by(decade) |> 
  summarise(missions = n_distinct(mission))

kable(decade_table, col.names = c("Decade", "Lunar missions"))
```

None in the 80s! Matches Wikipedia graph

### Landings

#### By country and outcome

```{r}
#| label: ggplot

# Make operator legend in order of frequency
country_counts <- table(moon_missions$country)
country_counts <- sort(country_counts, decreasing = TRUE) 
moon_missions$country <- factor(moon_missions$country, levels = names(country_counts))


moon_missions |> 
  filter(grepl("Lander", mission_type)) |> # Includes lander/launchers like LEM
  group_by(year) |> 
  arrange(desc(outcome_simple)) |> 
  mutate(y = row_number()) |> 
  ggplot(aes(x=year, y = y, text=paste0(
    mission_new,  
    "<br>Operator: ",operator,
    "<br>Launch date: ",launch_date, 
    "<br>Outcome: ", outcome,
    "<br><br>", description
  )
    )) +
  geom_point(aes(shape=outcome_simple, color=country), size=2.5) + # factors: alpha, colour, fill, shape, size 
  scale_shape_manual(values = c("Success" = 16, "Failure" = 4)) + # Circle for Success, Cross for Failure
  scale_color_manual(values=c('#377eb8','#e41a1c',"gold", '#4daf4a','#984ea3','#ff7f00','darkblue','#a65628','#f781bf','#999999')) +
  labs(colour="Country/Company", shape="Outcome") +
  ggtitle("All lunar landing attempts") +
  ylim(c(0,15)) + 
  theme_minimal() +
  theme(
    #panel.background = element_rect(fill="darkgrey", colour="darkgrey"),
    legend.position = "inside",
    legend.position.inside = c(.55, .55),
    legend.box.background = element_rect(color="black", linewidth=1),
    axis.title.y = element_blank(),    # Remove y-axis title
    axis.title.x = element_blank(),
    axis.text.y = element_blank(),     # Remove y-axis labels
    axis.ticks.y = element_blank(),    # Remove y-axis ticks
    axis.line.y = element_blank(),
    panel.grid.major.y = element_blank(),  # Remove y-axis major grid lines
    panel.grid.minor.y = element_blank()   # Remove y-axis minor grid lines
  )
```

#### Interactive

Can't colour and have useful legend at the same time here:

```{r plotly}

p <- moon_missions |> 
  filter(grepl("Lander", mission_type)) |> # Includes lander/launchers like LEM
  group_by(year) |> 
  arrange(desc(outcome_simple)) |> 
  mutate(y = row_number()) |> 
  ggplot(aes(x=year, y = y, text=paste0(
    mission_new,  
    "<br>Operator: ",operator,
    "<br>Launch date: ",launch_date, 
    "<br>Outcome: ", outcome,
    "<br><br>", description
  )
    )) +
  geom_point(aes(shape=outcome_simple)) + # factors: alpha, colour, fill, shape, size 
  scale_shape_manual(values = c("Success" = 16, "Failure" = 4)) + # Circle for Success, Cross for Failure
  ggtitle("All lunar landing attempts (hover over points for details)") +
  ylim(c(0,15)) + 
  theme_minimal() +
  theme(
    #legend.position="none",
    axis.title.y = element_blank(),    # Remove y-axis title
    axis.title.x = element_blank(),
    axis.text.y = element_blank(),     # Remove y-axis labels
    axis.ticks.y = element_blank(),    # Remove y-axis ticks
    axis.line.y = element_blank(),
    panel.grid.major.y = element_blank(),  # Remove y-axis major grid lines
    panel.grid.minor.y = element_blank()   # Remove y-axis minor grid lines
  )  +
  guides(color = guide_legend(title = NULL))  # Remove legend title

ggplotly(p, tooltip="text")  %>%
  layout(
    legend = list(
      title = "",  # Set legend title to an empty string to remove it
      bgcolor="lightgrey",
      x = 0.05,
      y=0.95
    )
  )

```

#### Locations

From Wikipedia: <https://en.wikipedia.org/wiki/File:Moon_landing_sites.svg>

Click to visit Wikipedia page for that mission

(Note: does not include most recent missions)

<img src="https://upload.wikimedia.org/wikipedia/commons/3/34/Moon_landing_sites.svg" alt="Markdown Logo"/>
